State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-09-12 analysis/paper/appendices/Appendix_STADEM_descrip.Rmd
incorported comments and revisions from Sarah Hoffman's review

Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations, trap catches and window counts, of the true number of fish crossing the dam in a state-space model [@royle2008hierarchical] with a weekly time-step. Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week. The log of the true number of fish crossing ($X_t$), is modeled as a random walk process [@Shumway2010].

@@

Thus, the unknown true number of fish crossing LGD each week, $X_t$, is estimated from two different data source: window counts and fish sampled in the trap. The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week. For weeks when we have a more precise estimate of the trap rate (i.e. weeks when lots of previously PIT tagged fish are crossing LGD), STADEM will tend to favor the estimate of total escapement based on the trap data, whereas when that trap rate is more uncertain (e.g. fewer PIT tagged fish to use in estimating the trap rate), STADEM will rely more on the window counts to estimate total escapement. During peak run times, when lots of fish are crossing LGD, estimates based on trap data and trap rates will be more precise, while estimates from the window counts may have more observation error due to so many fish passing the window. For weeks when the trap is down, STADEM relies exclusively on the window counts and nighttime passage data, but there will be more uncertainty in the estimates.

@@

There are two other processes that must be accounted for, first, the proportion of fish that cross the dam while the window is closed for counting (nighttime passage rate), and the second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted. Both rates can be estimated from previously PIT tagged fish that are crossing the dam each week.

>>>>

Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations (trap catches and window counts) of the true number of fish crossing the dam in a state-space model [@royle2008hierarchical] with a weekly time-step. Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week. The log of the true number of fish crossing ($X_t$), is modeled as a random walk process [@Shumway2010].

@@

Thus, the unknown true number of fish crossing LGD each week, $X_t$, is estimated from two different data source: window counts and fish sampled in the trap. The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week. For weeks when we have a more precise estimate of the trap rate (i.e. weeks when many previously PIT tagged fish are crossing LGD), STADEM will tend to favor the estimate of total escapement based on the trap data, whereas when that trap rate is more uncertain (e.g. fewer PIT tagged fish to use in estimating the trap rate), STADEM will rely more on the window counts to estimate total escapement. During peak run times, when many fish are crossing LGD, estimates based on trap data and trap rates will be more precise, while estimates from the window counts may have more observation error due to densely crowded (and visually obstructed) fish passing the window. For weeks when the trap is down, STADEM relies exclusively on the window counts and nighttime passage data, but there will be more uncertainty in the estimates.

@@

There are two other processes that must be accounted for: first, the proportion of fish that cross the dam while the window is closed for counting (nighttime passage rate), and second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted. Both rates can be estimated from previously PIT tagged fish that are crossing the dam each week.

### 2020-09-12 analysis/paper/appendices/Appendix_simulation_details.Rmd
incorporated suggestions from John Hargrove into manuscript
incorported comments and revisions from Sarah Hoffman's review

To examine the sensitivities of models to different fish behavior and dam operational scenarios, seven additional attributes are randomly assigned to each individual fish. Each attribute is randomly assigned a TRUE/FALSE using a Bernoulli $\mathcal{PDF}$ and a fixed probability parameter. Fish passage during the day-time (i.e., during periods of window operation) is modeled using one minus the night-time passage rate ($1 - \nu$). Window observations are conditioned on fish passing during the day and being observed at a set rate, $\gamma$. Whether fish $i$ is sampled by the adult trap is modeled on the weekly set trap rate, $\delta_t$. The rate of previously PIT-tagged fish is determined by $\lambda$, and their subsequent detection at the ladder PIT antenna is governed by $\kappa$. Fallback behavior is modeled with a common rate across all populations, $\psi$. Re-ascension occurs with probability $\rho$, conditioned on fish $i$ falling back. If fish $i$ falls back and re-ascends, the entire process described above is repeated, with some time-lag between initial ascension and re-ascension that is governed by a Poisson $\mathcal{PDF}$ with mean = 2 days. Fish may fall-back and re-ascend up to 3 times, allowing for the possibility of the same fish being counted or trapped multiple times.

>>>>

To model different fish behavior and dam operational scenarios, seven additional attributes are randomly assigned to each individual fish. Each attribute is randomly assigned a TRUE/FALSE using a Bernoulli $\mathcal{PDF}$ and a fixed probability parameter. Fish passage during the day-time (i.e., during periods of window operation) is modeled using one minus the night-time passage rate ($1 - \nu$). Window observations are conditioned on fish passing during the day and being observed at a set rate, $\gamma$. Whether fish $i$ is sampled by the adult trap is modeled on the weekly set trap rate, $\delta_t$. The rate of previously PIT-tagged fish is determined by $\lambda$, and their subsequent detection at the ladder PIT antenna is governed by $\kappa$. Fallback behavior is modeled with a common rate across all populations, $\psi$. Re-ascension occurs with probability $\rho$, conditioned on fish $i$ falling back. If fish $i$ falls back and re-ascends, the entire process described above is repeated, with some time-lag between initial ascension and re-ascension that is governed by a Poisson $\mathcal{PDF}$ with mean = 2 days. Fish may fallback and re-ascend up to 3 times, allowing for the possibility of the same fish being counted or trapped multiple times.
